Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-08-02 thesis/_book/_main.tex
Working on the presentation

\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
\author{Alicja Dobrzeniecka}
\date{2021-06-22}

@@

Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of data to learn from. This data originates from human activities and historical recordings such as texts, messages or speeches. It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies. There are many different types of models in NLP depending on a task that they are supposed to solve. However, all of them need as an input words represented by means of numbers and this is accomplished with word embeddings. Word embeddings are usually assigned the values based on the context in which the words appear. This means that the input data can have an enormous influence on the outcome. One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by, for example, modifying the word embeddings directly.

One can find various definitions trying to capture what bias and fairness actually are. With the choice of the definition, implications for the real-life applications may change as well. \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long). This indicates how complex the issue of bias is. Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias. \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described. By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against. Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them. Such discrimination takes place for example when one uses the information about someone's religious beliefs to evaluate whether they are competent for the work they apply for. In contrast, in indirect discrimination individuals appear to be treated equally, but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes. An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge. This job advert can discriminate indirectly against young people who can have the required skills, but not yet so much work experience. Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people. Finally, statistical discrimination consists in relying on group statistics to judge a person belonging to a given group. The topic of discrimination is entangled with another concept; fairness. It is essential to grasp some concepts of fairness to take them into consideration while designing implementation of some machine learning model. In \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that depending on the context and application different definitions may be applied.

There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models. \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focus on gender biases that may be observable in the representation of job occupations and gender in terms of numerical
values. The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
They also use analogy tasks to evaluate if a bias is present in a word embedding model. They check analogies by comparing pairs of word vectors. For example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots? First they subtract the word ``man'' from the word ``woman''\footnote{This is pointwise subtraction of vectors, but for the simplicity we will often talk about words while really meaning vectors. What is meant should be clear from the context.} and then
they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.

However, as \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are some limitations to this approach. In practice, most of analogies implementations do not return any input words. This means that it does not make sense to expect the algorithm to return the same profession for both woman and man. Therefore, this method of bias detection seems limited. Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination, as it is often subjective and without proper justification. Additionally, the choice of the parameter used in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
influences the results.

\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the topic of biases regarding race and gender. They apply knowledge from well-known psychological studies such as Implicit Association Test to investigate the relation between human stereotypical thinking and model-learnt biases to discover a close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).

\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a novel way of using cosine similarity to verify if the assumed resemblance between protected words and harmful attributes exists. They develop an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric. Their approach will be analyzed in details in the next chapter.

\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by measuring human reaction time for special tasks. Humans naturally exhibit some biases which do not always cause social concern. One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness. In general, people would rather associate flowers than insects with feeling pleasant, and this preference in a sense is a bias or
prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. This example is rather used to show that the methodology used in IAT makes sense. \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to obtain similar results to the ones that measured reaction time with the use of WEAT metric. Unfortunately, further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.

The most common bias detection methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way. One can find in this group methods such as Euclidean distance or cosine similarity (which is equivalent to dot product if the vectors are normalized). There are also other ways to detect the effects of biases. One can investigate the model performance on certain downstream tasks and validate if the model uses the gender or race information to complete the task. Unfortunately, the currently used methods employing cosine similarity, often aggregate the similarity values in a way that may lead to hasty conclusions. The averaging of values and the lack of attention to uncertainty involved in such aggregation may lead to an incomplete picture of the situation.

In the paper we indicate how current methods used to detect biases in natural language models are limited, if we investigate them from the perspective of Bayesian analysis. Our research enhances the current way in which the bias detection is performed to make sure that it is methodologically valid. The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.

@@

To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form. In the field of natural language processsing there are two main types of words representation --- localist and distributed. One-hot encoding is an example of a method used to achieve a localist representation of words. Here, each vector contains information only about a single data point. This is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1. An example of a localist representation is:

\begin{longtable}[]{@{}llllll@{}}
\toprule
word & 1 & 2 & 3 & 4 & 5 \\
\midrule
\endhead
woman & 1 & 0 & 0 & 0 & 0 \\
man & 0 & 1 & 0 & 0 & 0 \\
girl & 0 & 0 & 1 & 0 & 0 \\
boy & 0 & 0 & 0 & 1 & 0 \\
monarch & 0 & 0 & 0 & 0 & 1 \\
\bottomrule
\end{longtable}

In the example above it is clear that the length of the vectors increases with the number of words in a vocabulary. It is not a very computationally efficient representation. It has other flaws as well. For example, it is unable to capture the resemblance between words appearing in similar contexts.

In contrast to the localist representation, a distributed representation returns vectors that contain continuous values instead of discrete 1s and 0s. Word embeddings are a class of various techniques that allow one to represent words as distributed vectors. Such learned representations of text have certain properties. At least prima facie, they store similar (or at least co-occurring) words close to each other in a vector space. An example of distributed representation is:

\begin{longtable}[]{@{}lllll@{}}
\toprule
word & 1 & 2 & 3 & 4 \\
\midrule
\endhead
woman & 0.456 & 0.267 & 0.675 & 0.131 \\
man & 0.451 & 0.897 & 0.472 & 0.088 \\
girl & 0.604 & 0.262 & 0.414 & 0.706 \\
boy & 0.279 & 0.172 & 0.475 & 0.010 \\
monarch & 0.565 & 0.678 & 0.463 & 0.975 \\
\bottomrule
\end{longtable}

One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units. It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.

The numbers occurring in such representations are not random. They are learned in a process that uses a very shallow neural network. There are various types of techniques used for learning the vectors representations. One of the most straightforward ones is a skip-gram model.
Given a word, the model tries to predict its neighboring words from the sentence. The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words. The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus. One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict. After preparing such a fitted model, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.

Word embeddings have many applications in natural language processing. They are handy in document search and information retrieval. They also play their part in improving automatic translations. Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.

@@

Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected. Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.

It is worth mentioning one more point concerning cosine similarity. After the vectors are normalized to have length equal to 1, the inner product itself (often dot product) is used to measure the similarity, because it then equals the cosine similarity.

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}
Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted. Using this scale, one can compare vector similarities in a fairly clear manner. When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees). This tells us that the similarity between the vectors is small. As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity. It is also possible to obtain negative cosine similarity. If the value approaches -1, this intuitively means that the words are contrary to each other.

One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation. However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.

In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
\begin{align} \tag{Sim}
\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
 &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
\end{align}

The greater the similarity between two vectors, the smaller the distance between them. The cosine distance ranges between 0 and 2. If the vectors are in an opposite directions, the cosine distance is 2. And if the vectors are extremely similar then the cosine distance is very close to 0.

One should note that cosine distance is not exactly a distance measure, as it does not meet triangle inequality requirements. The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side. As shown in \href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{discussion in stats.stackexchange.com}\footnote{\url{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}} in the case of cosine distance it would have to fulfill this equation \(1+\mathsf{cos-sim}(A,C) < \mathsf{cos-sim}(A,B) + \mathsf{cos-sim}(B,C)\). If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.

@@

\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity between words as the outcome inner product of their normalized vectors.
They focus on examining what the geometry of a word embedding is in regard to ``he'' and ``she'' words. In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes. They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender. They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.

Here the gender bias of a word \(w\) is understood as its projection on the gender direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the gender direction is the top principal compontent of ten gender pair difference vectors). The underlying idea is that no bias is present if non-explicitly gender words are in equal distance to both elements in all explicitly gender pairs. Given the (ideally) gender neutral words \(N\) and the gender direction \(g\) the direct direct gender bias is defined as the average distance of the words in \(N\) from \(g\) (\(c\) is a parameter determining how strict we want to be):
\begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words. The plot below does not originate from the original paper but from \href{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}{kaggle.com R notebook using data from GloVe: Global Vectors for Word Representation}.\footnote{\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}.} However, similar visualization may be found there. Data used to create our plot is as follows.

Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}

Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
\normalsize

The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he). Inner product of two vectors expresses similarity between words. This assumption originates from the geometry and properties of a vector space. In the picture one may observe a correlation between the occupation and gender. Stereotypical male professions are closer to the ``he'' word and stereotypical female professions have greater similarity with ``she'' word. One could argue that this is a hint that there is some hidden bias information stored in the words vectors.

@@

\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a different approach towards finding similarities between classes of words. Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.

They modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper they report distances rather than similarities). Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and let each \(A_j\in A\) be a set of attributes stereotypically associated with a protected word). Then:
\begin{align}
S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
\end{align}

As the code has been provided by the authors we were able to reconstruct their results. The main steps in the procedure are as follows. Let us go through an example that refers to the process of hard debiasing on religious attributes.

First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset. The word embeddings have only 50 dimensions and the number of individual words from the dataset is 44895. The authors assume that the protected group should ideally not have high cosine similarity to stereotypical words. The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free. For instance, let's look at the religion-related words.

Protected words by religion type:

\begin{minipage}{1.0\textwidth}
\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}

\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}

\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
\end{minipage}

Stereotypical words by religion type:

\begin{minipage}{1.0\textwidth}
\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}

\textbf{"christian"       : ["judgemental", "conservative", "familial"]}

\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
\end{minipage}

We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown in Table 2.3

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllrrl}
\toprule
protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
\midrule
judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
\bottomrule
\end{tabular}}
\end{table}
\normalsize

\pagebreak

In the article there was no analysis of individual distances, but the general look at the means. The authors introduced a metric that tries to generalize WEAT in measuring the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race. In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype. They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in the case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam). Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.

In the article the authors also try to remove biases, as previously defined, from a word embedding. First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction. It is applied here to choose the subspace that contains the greatest amount of information. There can be many subspaces found in a given group. For example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:

\begin{minipage}{1.0\textwidth}
\textbf{["judaism", "christianity", "islam"]}

\textbf{["jew", "christian", "muslim"]}

\textbf{["synagogue", "church", "mosque"]}
\end{minipage}

The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words. This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed. In some cases this knowledge is useful, but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.

After finding the bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed. In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and stereotype is smaller.

In the final step they evaluate the results. The cosine distances are calculated again but this time using the debiased vocabulary. After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning. If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved. As the cosine distance increases, it is assumed that the association between protected and stereotypical words decreases.

@@

\textbf{1. Selection of attributes}

The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that the list is rather uneven.

There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset. There are however some ways to estimate sample sizes needed to ensure that the results are meaningful if the effect is present. Our research will show that the numbers used
are rather insufficient.

\textbf{2. No attention to distributions and details}

The authors use the mean of mean average cosine similarities to measure multi-class similarity between protected word and harmful stereotypes. They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class. If one takes closer look at the individual values that are taken for the calculations it turns out that there is a bunch of outliers and surprisingly dissimilar words. We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we obtain a proper insight into the data.

\textbf{3. Hiding the impact of uncertainty}

With such a method the uncertainty involved is not really considered which makes it even more difficult to give reasonable interpretations of the results. We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.

\pagebreak

\textbf{4. No word class distinction and no control group}

In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one. Not all of the stereotypical words have to be considered as harmful for all of the religions. One should investigate the religions separately as some of them may have stronger harmful associations that others. One should also include control groups to have a way of comparing the stereotypical results with neutral or human-like words. Later in the text we will explain in details reasons for introducing control groups. In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.

\textbf{5. Interpreting the results}

Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation. \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the averages of cosine distance per group (gender, race, religion). For now let us focus now on analyzing the values relating to religious biases. Here is the relevant fragment of table:

\begin{longtable}[]{@{}ll@{}}
\toprule
Religion Debiasing & MAC \\
\midrule
\endhead
Biased & 0.859 \\
Hard Debiased & 0.934 \\
Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
\bottomrule
\end{longtable}

MAC stand for mean average cosine similarity, although in reality the the table contains mean distances. What may attract attention is the fact that the value of cosine distance in ``Biased'' category is already quite high even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias. However the authors aim to mitigate ``biases'' in vectors with such great distance to make it even larger. Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful. This is in general the problem of scale and the lack of universal intervals. In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random. We will use highest posterior density intervals, in line with Bayesian methodology.

\textbf{6. The curse of dimensionality}

In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space. According to the article
\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Curse of dimensionality at analyticsindiamag.com} as the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms. One may notice that more data should contribute to greater amount of information, but more information also means greater risk of noise and distractions in data. At the same time, modern solutions are often adapted to smaller dimensions and their results in higher ones are not intuitive, or may be prone to error.

Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality. According to \protect\hyperlink{ref-Venkat2018Curse}{Venkat} (\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to consider this phenomenon when searching for word similarities in higher dimensions. An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase. First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation. The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data. Yet, according to some articles cosine similarity in high dimensions is not reliable enough as it may be the case that choosing words at random may result in similar values as when picking them consciously.

\begin{figure}
\centering
\includegraphics{../images/curseOfDimensionality.png}
\caption{curse of dimensionality, number of dimensions on the x axis, standard deviation of similarity on the y axis}
\end{figure}

@@

@@

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
\normalsize

@@

@@

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
\normalsize

@@

We propose the use of Bayesian methods to measure uncertainty in bias detection. There are a few advantages of this method. One of them is that including uncertainty enables one to directly observe the influence of sample sizes. Analyzing individual words and connection coefficients, one may notice how \texttt{neutral} words have smaller uncertainty intervals and \texttt{different} or \texttt{associated} quite the opposite. One of the reasons for such an outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) article. In our approach, we also pay attention to the distribution and details regarding anomalous values. With the use of simple visualizations that we introduced before, we were able to indicate suspicious cosine distance values. Additionally, we compare in details how the cosine distance values and uncertainty change after the debiasing. One can investigate then how the individual vectors changed and if it is what was expected. Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.
We created summary tables for each of the datasets: Google word embeddings (Table \ref{tab:dTabG}), Reddit (Table \ref{tab:dTabR}), and Reddit Debiased (Table \ref{tab:dTabRD}).

\vspace{1mm}
\footnotesize
\begin{table}

\caption{\label{tab:dTabG}Overview statistics  (GoogleNews embedding)}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llll}
\toprule
  & Religion (percent) & Gender (percent) & Race (percent)\\
\midrule
A close to N & 66.67 & 14.29 & 80\\
A close to H & 66.67 & 21.43 & 100\\
H closer than A & 20 & 21.43 & 50\\
D similar with A & 93.33 & 42.86 & 100\\
D closer than A & 33.33 & 21.43 & 30\\
General HPDI outside of  0 & no & no & no\\
General D similar to A & slightly not & slightly not & slightly not\\
\bottomrule
\end{tabular}}
\end{table}
\normalsize

\vspace{1mm}
\footnotesize
\begin{table}

\caption{\label{tab:dTabR}Overview statistics  (Reddit embedding)}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llll}
\toprule
  & Religion (percent) & Gender (percent) & Race (percent)\\
\midrule
A close to N & 66.67 & 0 & 40\\
A close to H & 60 & 21.43 & 50\\
H closer than A & 33.33 & 0 & 0\\
D similar with A & 100 & 92.86 & 90\\
D closer than A & 33.33 & 42.86 & 30\\
General HPDI outside of 0 & no & yes & no\\
General D similar to A & slightly not & yes & slightly not\\
\bottomrule
\end{tabular}}
\end{table}
\normalsize

\vspace{1mm}
\footnotesize
\begin{table}

\caption{\label{tab:dTabRD}Overview statistics  (Reddit embedding, debiased)}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llll}
\toprule
  & Religion (percent) & Gender (percent) & Race (percent)\\
\midrule
A close to N & 73.33 & 7.14 & 50\\
A close to H & 46.67 & 21.43 & 70\\
H closer than A & 66.67 & 0 & 20\\
D similar with A & 100 & 100 & 80\\
D closer than A & 66.67 & 50 & 30\\
General HPDI outside of 0 & no & yes & no\\
General D similar to A & yes & yes & yes\\
General H moved closer to A & yes & no & no\\
General N moved closer & no & yes & yes\\
\bottomrule
\end{tabular}}
\end{table}
\normalsize

Let us first analyse the general observations from estimated coefficients mean introduced in Section \ref{sec:datasetsLevelCoeffs} on dataset-level coefficients. For Google embeddings the HPDIs for all class coefficients (associated, different, human, and none) include zero. This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar. This indicates how including the uncertainty may change the use and interpretation of \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) MAC metric. It seems that if one focuses only on differences between means of means, this is too simplistic. In case of Reddit word embeddings the situation is similar although HPDI is below 0 for Gender class when looking at \texttt{associated} and \texttt{different} mean coefficient. This can suggest that there is indeed slightly stronger impact of these attributes. One should also notice how in general \texttt{associated} and \texttt{different} coefficients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1. This suggests that again the impact of associated attributes and different ones is not clear at first sight. Finally, let's compare the HPDIs for Reddit and Reddit debiased datasets. For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero. However for Gender dataset there is no significant change. This is of course a general look at the group coefficients. Let's now analyze the individual words.

In Reddit table one can observe that for Religion the cosine distance results for the associated attributes are for approximately 60\% of the words close to human attributes as well. This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently of whether they are stereotypical or neutral-human words. One should also pay attention to the fact that for all of the \texttt{associated} and \texttt{different} attributes in Religion, the uncertainty intervals overlap at some point. What is even more surprising is that for protected words, such as ``torah'' the \texttt{associated} attribute has the the cosine distance slightly over 1, which means no positive similarity! If the protected words that we chose do not have high similarity with harmful associated attributes, then one should consider at least three scenarios. The first one is that the choice of the protected words and attributes may be corrupted. The second one is that the metric is unable to catch the hidden bias properly. The third one is that there is actually no bias between the words. Regardless of which scenario one considers, it is essential to take a look at the individual values before averaging them or aggregating in other ways. It seems that using Bayesian method can enhance the process of verifying the hypotheses concerning the choice of protected words and attributes.

>>>>

\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
\author{Alicja Dobrzeniecka}
\date{2021-08-02}

@@

Placeholder

@@

@@

@@

@@

@@

@@

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}
\normalsize

@@

@@

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}
\normalsize

@@

@@

Placeholder
